Harmony: A Systems Programming Language and Model Checker

### 2022-01-19 paper/paper.tex
Improved description of "breakable" instructions

\begin{compactitem}
\item \emph{basic operations}: a deterministic operation that only
involves a single context and no shared state, such as pushing a value
onto the stack or increment the program counter.
\item \texttt{Frame/Return}: method start/end.
\item \texttt{Choose}: nondetermistic selection from a set.
\item \texttt{Apply}: method call \emph{or} index into a dictionary.
\item \texttt{Spawn}: thread creation.
\item \texttt{AtomicInc/AtomicDec}: atomic section enter/leave.
\item \texttt{SetIntLevel}: interrupt enable/disable.
\item \texttt{Stop/Go}: continuation save/restart.
\item \texttt{Assert}: value check.
\item \texttt{Print}: output.
\end{compactitem}

@@

There are three sources of nondetermism in Harmony: \texttt{Choose}
operations, thread interleaving, and interrupts.
When it comes to interleaving, an important observation is that
local operations on the state of a context (updates to its program
counter, local variables, or stack) are not visible to other threads.
An important optimization that significantly
reduces the size of the set of generated states is to treat a
sequence of operations as atomic if only its first operation is an
operation that accesses the shared state (\texttt{Load}, \texttt{Store},
\texttt{Del}, or \texttt{Stop}) or a \texttt{Choose} or \texttt{Print}
operation.  We call such operations \emph{breakable}.
Note that this optimization is essentially Partial Order
Reduction~\cite{Val91}.

@@

Given a state and a context, Charm starts with applying the instruction
pointed to by the context's program counter to the state.
Then it continues to apply instructions as long as
the active set is the singleton set (such as when the thread is in
atomic mode) or, if not, as long as the next instruction is not
breakable.  Only the resulting state is added to the Kripke structure,
and the edge indicates the starting context.  During this loop, Charm
does check if the same state recurs---if so, it aborts
and reports an infinite loop.

@@

In practice, about half of the work of the model checking is trivially
parallelizable, allowing for significant speedups on multi-core machines.

>>>>

\begin{compactitem}
\item \emph{local operations}: a deterministic operation that only
involves a single context and no shared state, such as pushing a value
onto the stack or increment the program counter.
\item \texttt{Load/Store/Del}: an operation that modifies the shared state.
\item \texttt{Print}: output.
\item \texttt{Choose}: nondetermistic selection from a set.
\item \texttt{Frame/Return}: method start/end.
\item \texttt{Apply}: method call \emph{or} index into a dictionary.
\item \texttt{Spawn}: thread creation.
\item \texttt{AtomicInc/AtomicDec}: atomic section enter/leave.
\item \texttt{SetIntLevel}: interrupt enable/disable.
\item \texttt{Stop/Go}: continuation save/restart.
\item \texttt{Assert}: value check.
\end{compactitem}

@@

There are three sources of nondetermism in Harmony: \texttt{Choose}
operations, thread interleaving, and interrupts.
Each of these cause branching in the state graph and contribute
to the combinatorial explosion of states.
When it comes to interleaving, an important observation is that
local operations on the state of a context (updates to its program
counter, local variables, or stack) are not visible to other threads
and therefore two such operations executed by different threads
commute.
Operations that do not commute that way are Load, Store, and Del
instructions, as well as the Print instruction.  We call those
instructions \emph{breakable}.
An important optimization that significantly
reduces the size of the set of generated states is to treat a
sequence of operations as atomic if only its first instruction
is breakable.  Given two such sequences $\sigma_1$ and $\sigma_2$,
the execution of any interleaving would either be equivalent to
$\sigma_1\sigma_2$ or $\sigma_2\sigma_1$.
This is another example of Partial Order Reduction~\cite{Val91}.

@@

Given a state and a context, Charm starts with applying the instruction
pointed to by the context's program counter to the state.
Then it continues to apply instructions as long as
the active set is the singleton set (such as when the thread is in
atomic mode) or, if not, as long as the next instruction is not
breakable.  Only the resulting state is added to the Kripke structure,
and the resulting edge is labeled with the starting context.
During this loop, Charm
does check if the same state recurs---if so, it aborts
and reports an infinite loop.

@@

% In practice, about half of the work of the model checking is trivially
% parallelizable, allowing for significant speedups on multi-core machines.
